---
title: "STAC51_A3"
author: "Rafsaan Sanvir"
date: "2024-04-02"
output: word_document
---
1a)
```{r}
data1 = data.frame(
  X = c(5, 3, 4, 5, 4, 6, 5, 3, 6, 5),
  Y = c(53, 34, 45, 49, 45, 77, 62, 37, 75, 54)
)

model1 = glm(formula = Y~X, data = data1, family = poisson(link = log))
summary(model1)
```
estimate of model in written sheet 

1b)
```{r}
full_model = model1 
reduced_model = glm(formula = Y~1, data = data1, family = poisson(link = log))
lr_test = anova(reduced_model, full_model, test = "Chisq")
print(lr_test)
```
Since p value extremely small, less than alpha = 0.05, reject null hypothesis. 
Meaning explanatory variable X has significant effect on response variable Y. 

1c) 
```{r}
saturated_MLE = prod(dpois(data1$Y, lambda = data1$Y))
#note: mean for saturated model is yi/ni, but ni=1 so mean just yi,  
cat("saturated MLE estimate is:" , saturated_MLE, "\n")
```

1d) 
```{r}
alpha = coefficients(model1)[1]
beta = coefficients(model1)[2]
mean_xi = exp(alpha + beta * data1$X)
full_model_MLE = prod(dpois(data1$Y, lambda = mean_xi))
cat("full model MLE estimate is:" , full_model_MLE, "\n")
```
1e) 
```{r}
mean_xi2 = exp(alpha)
reduced_model_MLE = prod(dpois(data1$Y, lambda = mean_xi2)) 
print(reduced_model_MLE)
cat("reduced model MLE estimate is:" , reduced_model_MLE, "\n")
```
1f) (not complete)
```{r}
null_di = -2 * log(saturated_MLE / reduced_model_MLE)
residual_di = -2 * log(full_model_MLE / reduced_model_MLE)
cat("null deviance is ", null_di, "\n")
cat("residual deviance is ", residual_di, "\n")

```


2c) 
```{r}
library(Rsolnp)
y = c(11, 14, 24, 22, 29) 
ll = function(pi) {
  ll = log(dmultinom(y, prob = c(pi[1], pi[2], pi[3], pi[4], pi[5])))
  # return the negative to maximize rather than minimize
  return(-ll)
}
eqn = function(pi) {
  z1 <- pi[1]-pi[2] # For pi1+pi2+pi3 =1
  z2 <- pi[3] -pi[4] # For pi1=2pi2 or pi1-2pi2 =0
  return(c(z1, z2))
}
constraints = c(0,0)
lpi = c(0,0,0,0,0)
upi = c(1,1,1,1,1)
pi0 = c(1/5, 1/5, 1/5, 1/5, 1/5)
ctrl = list(trace = 0)
sol1 = solnp(pi0, fun = ll, eqfun = eqn, eqB = constraints, LB=lpi,
             UB=upi, control=ctrl)
cat("MLe estimate is: ", sol1$pars, "\n")
```

2d) 
```{r}
n = sum(y)
muhat = sol1$pars * n 
LRT_2d = sum(((y-muhat)^2)/muhat)
p_value_LRT = 1-pchisq(LRT_2d, df = length(y) - 1)
cat("p value is: ", p_value_LRT, "\n")
```
At alpha = 0.05 level of significance, we fail to reject null hypothesis since 
p value is greater 


```{r}
gsq_2e = 2 * sum( y * log(y/muhat))
p_value_gsq = 1-pchisq(gsq_2e, df = length(y) - 1)
cat("p value is: ", p_value_gsq, "\n")
```
At alpha = 0.05 level of significance, we fail to reject null hypothesis since 
p value is greater 
 
 
3a) 
```{r}
N = 100
alpha = 0.05
pi0 = c(0.15, 0.15, 0.3, 0.4)
pi = c(0.1, 0.2, 0.3, 0.4)
crt_value1 = qchisq(1-alpha, df = length(pi0) - 1)
lambdaP = N * sum(((pi-pi0)^2)/pi0)
power_P = 1-pchisq(crt_value1, 3, lambdaP)
cat("power is ", power_P)
```
3b)
```{r}
lambdaLRT = 2*N*sum(pi*log(pi/pi0))
power_LRT = 1-pchisq(crt_value1, 3, lambdaLRT)
cat("power is: ", power_LRT)

```
3c) 
```{r}
N=0
power = 0.05 
while(power <0.9){
  N=N+1
  lambdaP = N * sum(((pi-pi0)^2)/pi0)
  power = 1-pchisq(crt_value1, 3, lambdaP)
}
cat("at N =", N, ", we get a power of ", power)
```
3d) 
```{r}
N2=0
power2 = 0.05 
while(power2 <0.9){
  N2=N2+1
  lambdaLRT = 2*N2*sum(pi*log(pi/pi0))
  power2 = 1-pchisq(crt_value1, 3, lambdaLRT)
}
cat("at N =", N2, ", we get a power of ", power2)
```


4a) 
```{r}

data2 = data.frame(
  DoseLevel = c(0, 1, 2, 3, 4, 5),  # Convert to factor
  DeadMoths = c(1, 4, 9, 13, 18, 20)
)
#probability of death for x moths dead is x/20
model2 = glm(cbind(DeadMoths, 20 - DeadMoths) ~ DoseLevel, data = data2, family = binomial)
summary(model2)
```
logistic regression model written on paper using values 

4b)
```{r}
null_model <- glm(cbind(DeadMoths, 20 - DeadMoths) ~ 1, data = data2, family = binomial)
summary(null_model)
lrt_test_4b = anova(null_model, model2, test = "Chisq")
lrt_test_4b
```
p value much smaller than even 0.01, so reject null hypothesis, meaning 
doselevel has significant effect on number of mothsn dead. 

4c) 
```{r}
probability_death = model2$coefficients[1] + model2$coefficients[2]*3
cat("probability od death at dose level 3 is: ", probability_death)
```

4d) El0.75 is the dose level at which the death level of moths reach 75%, predicted by the full regression model. 
```{r}
lethal_dose = (log(0.75/(1-0.75)))-model2$coefficients[1]/model2$coefficients[2]
cat("at a dose level of ", lethal_dose, "75% of the moths are expected to die")
```

4e) 
```{r}
pearson_residual = resid(model2, type = "pearson")
hat_values = hatvalues(model2)
std_pearson_residual = pearson_residual /sqrt(1-hat_values)
cat("standardized pearson residual values are: ", std_pearson_residual)
```

4f) 
```{r}
pearson_chi_sq = sum(pearson_residual^2)
p_value = pchisq(pearson_chi_sq, df = length(pearson_residual) -model2$rank, lower.tail = FALSE)
p_value
```
since p value much greater than even 0.8 (normally we use a = 0.05), we fail to 
reject null hypothesis which means that the full model provides a good fit 
for the data. 

